Unlocking Sensitivity for Visibility-based Estimators of the 21\,cm Reionization Power Spectrum

### 2017-10-13 Rotation paper/draft_working.tex
correct typo

Analysis pipelines for the 21\,cm power spectrum typically fall into two categories. In the first, images are formed through rotation synthesis, and after a foreground mitigation step, are Fourier transformed to construct a power spectrum \citep{MWAresult0, MWAresult1, LOFARresult}. An alternative technique works directly with visibilities from baselines, delay-transforming and cross-multiplying them to form the power spectrum \citep{delay-transform}. This technique avoids many systematics associated with combining data from different baselines and tracks the native sampling of the interferometer. An example of the visibility-based pipeline is presented in \cite{Ali2015}, which provides power spectrum
measurements with the 64-element version of PAPER (henceforth as PAPER-64). However, one disadvantage of existing visibility-based pipelines is their lack of treatment of partial redundancy. Baselines that are slightly different in length and orientation
``rotate into'' each other at a time offset. Baselines of different lengths and orientations therefore contain partially coherent information. This is the basis of earth-rotation synthesis. While imaging based power-spectrum pipelines naturally include all redundancy information, visibility-based pipelines so far only cross multiply fully redundant baselines, i.e., baselines of the same length and orientation.

@@

We shall use the 128-element PAPER array (henceforth referred to as PAPER-128) to motivate our formalism and demonstrate our method, and extend our results to several HERA configurations in Section \ref{sec:arrconf}. 
The PAPER array is located in the Karoo desert in South Africa (30:43:17.5
S, 21:25:41.8 E). The layout pattern is shown
in Fig. \ref{fig:AntPos}. The array consists of a 112-element core in a rectangular grid, and 16 ``outriggers'' used primarily to aid calibration.  In the bottom panel, the two baselines marked in red are an example of an equivalent pair. We denote a equivalency class of baselines in the PAPER grid by their separations, in this case  \{2,0\}, for the
antennas are separated by 2 units east and 0 units north. Similarly,
the baselines marked in blue and yellow are respectively examples
of \{2,1\} and $\{2, -1\}$.
Note that \{2,0\} and $\{-2, 0\}$ are the same class of baselines and should not be included twice in power-spectrum measurements. Antennas in purely north-south baselines
are close (4m), and hence these baselines are not suitable
for sensitive measurements due to cross-coupling. On the other hand, the small North-South separation means that classes such as \{2,0\} and  \{2,1\} are expected to be near-equivalent. The PAPER-64 analysis of \cite{Ali2015} uses three classes of baselines, the PAPER-128
equivalent of which are 
\{2, 0\}, \{2, 1\} and $\{2, -1\}$. There, baselines are cross-multiplied within each class. This paper provides the method for inter-class multiplications. We will use the short hand notation $\{m,n:m',n'\}$ to denote a pair of baseline classes to be cross-multiplied.

@@

A delay-transformed visibility $V(\mathbf{b}, \tau) $ is related to the spherical Fourier-Bessel modes by
\begin{equation}
V(\mathbf{b}, \tau) = \sqrt{\frac{2}{\pi}} \sum_{\ell m} \int \! dk\, k^2 g_{\ell m} (k; \mathbf{b}, \tau) \overline{T}_{\ell m} (k),
\end{equation}
where
\begin{equation}
g_{\ell m} (k ; \mathbf{b}, \tau) = \int \! d\Omega d\nu B(\hat{\mathbf{r}},\nu) Y_{\ell m} (\hat{\mathbf{r}}) j_\ell (k r) e^{i 2 \pi \nu (\tau - \mathbf{b}\cdot \hat{\mathbf{r}} / c)}
\end{equation}
and the analogous quantity for $V_\psi$ is given by
\begin{eqnarray}
\label{eq:glmpsi}
g_{\ell m}^\psi (k ; \mathbf{b}, \tau) = \int && d\Omega d\nu B(\boldsymbol \Gamma \hat{\mathbf{r}},\nu) Y_{\ell m} (\hat{\mathbf{r}}) j_\ell (k r) \nonumber \\
&& \times  e^{i 2 \pi \nu (\tau - \mathbf{b}\cdot \boldsymbol \Gamma \hat{\mathbf{r}} / c) + i \psi_\nu}.
\end{eqnarray}
Note that unlike the primary beam and the fringe pattern, the spherical harmonic function does \emph{not} rotate, since it originated from a spherical harmonic expansion of the sky temperature, which is fixed. With these expressions and a little algebra combining Eqs. \eqref{eq:SHB} through \eqref{eq:glmpsi}, we arrive at a form of the two-baseline estimator (compare with Eq. \eqref{eq:final}):
\begin{eqnarray}
\hat{P}(k) &=& \frac{ \langle V^{*}(\boldsymbol{b},\tau)V_{\psi}(\boldsymbol{b'},\tau) \rangle}{\Theta} \nonumber \\
&=& \sum_\ell \int \!dk\, W_{\ell}(k; \mathbf{b}, \mathbf{b}^\prime , \tau) P(k),
\end{eqnarray}
where we have defined the \emph{window function}
\begin{equation}
W_{\ell}(k; \mathbf{b}, \mathbf{b}^\prime , \tau) \equiv \frac{2 k^2}{\pi \Theta} \sum_m g_{\ell m} (k ; \mathbf{b}, \tau) g_{\ell m}^{\psi*} (k ; \mathbf{b}^\prime, \tau).
\end{equation}
By construction, the window function sums to unity when integrated over all $k$ and summed over all $\ell$. We may therefore interpret our estimator $\hat{P}$ of the power spectrum to be a weighted average over all possible $\ell$ and $k$ modes, with the window function providing the weights of this average. This means that window functions can serve as indicators of foreground leakage into the EoR window. One writes down the estimator for a hypothetical measurement of a power spectrum mode within the EoR window, and additionally computes the window function for that estimator. Ideally, the window function for $\hat{P}(k)$ will be sharply peaked around $k$. In general, however, window functions will have wings that encroach upon other modes. If the window function is substantially non-zero at low-$k$ modes---which is where foreground emission typically resides, given its spectral smoothness---then the EoR window will be contaminated by foregrounds.

@@

First we can average the visibilities of the equivalent baselines. Since the core of PAPER-128 is a 16 by 7 antenna configuration, there
are $M\equiv(16-|m|)\times(7-|n|)$ copies of the baseline class $\{m,n\}$. This means
that if we add visibility measurements of all the equivalent baselines,
we get a factor of $\sqrt{M}$ reduction in the noise
level $\sigma_N$ of the visibility. The sensitivity contribution of $\{m,n\}$, cross multiplied with $\{m',n'\}$  thus roughly scales as $\sqrt{(16-|m|)(7-|n|)(16-|m'|)(7-|n'|)}=\sqrt{MM'}$.
For cross-multiplications of nearly equivalent baselines of
types $\{m,n\}$ and $\{m',n'\}$, we get an effective weight:

@@

\caption{Pairplots of the top contributing baseline pairs in three arrays. Plotted properties are optimal time offset $\Delta t_{\boldsymbol{bb'}}$ (in hours),  effective weight $\widetilde{\Theta}_{\boldsymbol{bb'}}$, and baseline length $\bar{b}$ (in meters). Only those points with $\widetilde{\Theta}_{\boldsymbol{bb'}}>0.01$ are shown (the weight for the top class of equivalent pair is normalized to 1). Scatter points of the three arrays overlap, in the order indicated by the legend. }
\label{fig:pairplot}
\end{figure}

In Fig. \ref{fig:pairplot} we present pair-distributions of three different properties for baselines that contribute well to the sensitivity ($\widetilde{\Theta}_{\boldsymbol{bb'}}>0.01$, where again the weight for the top class of equivalent pair is normalized to 1). The properties shown are effective weight $\widetilde{\Theta}_{\boldsymbol{bb'}}$, optimal time offset $\Delta t_{\boldsymbol{bb'}}$, and average baseline length $\bar{b}$. We only show three of the mentioned arrays for visual clarity. The other two results are qualitatively similar. We point out some key features:
\begin{itemize}

@@

The dashed lines represent the values when only the equivalent baseline-pairs are used. The small and unfractured HERA-37, with no gap (like in HERA-350) or short nearly equivalent baselines (like in PAPER 128), will not benefit much from the nearly equivalent baselines. Near-equivalent baselines make an increasingly significant contribution to the total sensitivity as the number of array elements are increased. PAPER-128 is designed with highly redundant nearly equivalent baselines, and thus  these baselines start contributing at higher $\widetilde{\Theta}_{min}$. The fractured HERA configurations will benefit even more from nearly equivalent baselines at low $\widetilde{\Theta}_{min}$ due to the presence of more classes of such pairs. The stepwise pattern seen in almost all cases are characteristic of a regular grid; as we step to lower $\widetilde{\Theta}_{\boldsymbol{bb'}}$ large groups of baseline pair classes get included in ``batch''. Thus the ``unfractured'' HERA-37, with more regularity in its antenna configurations than the larger counterparts, exhibit more step patterns in fractional sensitivity. 
%\begin{centered}
\begin{figure}[H]
\includegraphics[width=\linewidth]{osens}
\caption{Fractional sensitivity of redundant arrays as a function of the minimum effective weight. Dashed lines represent when only equivalent baseline pairs are used, while solid lines indicate use of both the equivalent and nearly equivalent baselines are used. The vertical axis is normalized independently for each array to the total sensitivity when all baseline pairs, equivalent and near-equivalent, are used.}
\label{fig:osens}
\end{figure}
%\end{centered}

>>>>

Analysis pipelines for the 21\,cm power spectrum typically fall into two categories. In the first, images are formed through rotation synthesis, and after a foreground mitigation step, are Fourier transformed to construct a power spectrum \citep{MWAresult0, MWAresult1, LOFARresult}. An alternative technique works directly with visibilities from baselines, delay-transforming and cross-multiplying them to form the power spectrum \citep{delay-transform}. This technique avoids many systematics associated with combining data from different baselines and tracks the native sampling of the interferometer. An example of the visibility-based pipeline was presented in \cite{Ali2015}, which provided power spectrum
measurements with the 64-element version of PAPER (henceforth as PAPER-64). However, one disadvantage of existing visibility-based pipelines is their lack of treatment of partial redundancy. Baselines that are slightly different in length and orientation
``rotate into'' each other at a time offset. Baselines of different lengths and orientations therefore contain partially coherent information. This is the basis of earth-rotation synthesis. While imaging based power-spectrum pipelines naturally include all redundancy information, visibility-based pipelines so far only cross multiply fully redundant baselines, i.e., baselines of the same length and orientation.

@@

We shall use the 128-element PAPER array (henceforth referred to as PAPER-128) to motivate our formalism and demonstrate our method, and extend our results to several HERA configurations in Section \ref{sec:arrconf}. 
The PAPER array \textbf{was} located in the Karoo desert in South Africa (30:43:17.5
S, 21:25:41.8 E). The layout pattern is shown
in Fig. \ref{fig:AntPos}. The array \textbf{consisted} of a 112-element core in a rectangular grid, and 16 ``outriggers'' used primarily to aid calibration.  In the bottom panel, the two baselines marked in red are an example of an equivalent pair. We denote a equivalency class of baselines in the PAPER grid by their separations, in this case  \{2,0\}, for the
antennas are separated by 2 units east and 0 units north. Similarly,
the baselines marked in blue and yellow are respectively examples
of \{2,1\} and $\{2, -1\}$.
Note that \{2,0\} and $\{-2, 0\}$ are the same class of baselines and should not be included twice in power-spectrum measurements. Antennas in purely north-south baselines
are close (4m), and hence these baselines are not suitable
for sensitive measurements due to cross-coupling. On the other hand, the small North-South separation means that classes such as \{2,0\} and  \{2,1\} are expected to be near-equivalent. The PAPER-64 analysis of \cite{Ali2015} \textbf{used} three classes of baselines, the PAPER-128
equivalent of which \textbf{were} 
\{2, 0\}, \{2, 1\} and $\{2, -1\}$. There, baselines \textbf{were} cross-multiplied within each class. This paper provides the method for inter-class multiplications. We will use the short hand notation $\{m,n:m',n'\}$ to denote a pair of baseline classes to be cross-multiplied.

@@

A delay-transformed visibility $V(\mathbf{b}, \tau) $ is related to the spherical Fourier-Bessel modes by
\begin{equation}
V(\mathbf{b}, \tau) = \sqrt{\frac{2}{\pi}} \sum_{\ell m} \int \! dk\, k^2 g_{\ell m} (k; \mathbf{b}, \tau) \overline{T}_{\ell m} (k),
\end{equation}
where
\begin{equation}
g_{\ell m} (k ; \mathbf{b}, \tau) = \int \! d\Omega d\nu B(\hat{\mathbf{r}},\nu) Y_{\ell m} (\hat{\mathbf{r}}) j_\ell (k r) e^{i 2 \pi \nu (\tau - \mathbf{b}\cdot \hat{\mathbf{r}} / c)}
\end{equation}
and the analogous quantity for $V_\psi$ is given by
\begin{eqnarray}
\label{eq:glmpsi}
g_{\ell m}^\psi (k ; \mathbf{b}, \tau) = \int && d\Omega d\nu B(\boldsymbol \Gamma \hat{\mathbf{r}},\nu) Y_{\ell m} (\hat{\mathbf{r}}) j_\ell (k r) \nonumber \\
&& \times  e^{i 2 \pi \nu (\tau - \mathbf{b}\cdot \boldsymbol \Gamma \hat{\mathbf{r}} / c) + i \psi_\nu}.
\end{eqnarray}
Note that unlike the primary beam and the fringe pattern, the spherical harmonic function does \emph{not} rotate, since it originated from a spherical harmonic expansion of the sky temperature, which is fixed. With these expressions and a little algebra combining Eqs. \eqref{eq:SHB} through \eqref{eq:glmpsi}, we arrive at a form of the two-baseline estimator (compare with Eq. \eqref{eq:final}):
\begin{eqnarray}
\hat{P}(k_{\bar{b}, \tau}) &=& \frac{ \langle V^{*}(\boldsymbol{b},\tau)V_{\psi}(\boldsymbol{b'},\tau) \rangle}{\Theta} \nonumber \\
&=& \sum_\ell \int \!dk\, W_{\ell}(k; \mathbf{b}, \mathbf{b}^\prime , \tau) P(k),
\end{eqnarray}
where we have defined the \emph{window function}
\begin{equation}
W_{\ell}(k; \mathbf{b}, \mathbf{b}^\prime , \tau) \equiv \frac{2 k^2}{\pi \Theta} \sum_m g_{\ell m} (k ; \mathbf{b}, \tau) g_{\ell m}^{\psi*} (k ; \mathbf{b}^\prime, \tau).
\end{equation}
By construction, the window function sums to unity when integrated over all $k$ and summed over all $\ell$. We may therefore interpret our estimator $\hat{P}$ of the power spectrum to be a weighted average over all possible $\ell$ and $k$ modes, with the window function providing the weights of this average. This means that window functions can serve as indicators of foreground leakage into the EoR window. One writes down the estimator for a hypothetical measurement of a power spectrum mode within the EoR window, and additionally computes the window function for that estimator. Ideally, the window function for $\hat{P}(k)$ will be sharply peaked around $k$. In general, however, window functions will have wings that encroach upon other modes. If the window function is substantially non-zero at low-$k$ modes---which is where foreground emission typically resides, given its spectral smoothness---then the EoR window will be contaminated by foregrounds.

@@

First we can average the visibilities of the equivalent baselines. Since the core of PAPER-128 was a 16 by 7 antenna configuration, there
are $M\equiv(16-|m|)\times(7-|n|)$ copies of the baseline class $\{m,n\}$. This means
that if we add visibility measurements of all the equivalent baselines,
we get a factor of $\sqrt{M}$ reduction in the noise
level $\sigma_N$ of the visibility. The sensitivity contribution of $\{m,n\}$, cross multiplied with $\{m',n'\}$  thus roughly scales as $\sqrt{(16-|m|)(7-|n|)(16-|m'|)(7-|n'|)}=\sqrt{MM'}$.
For cross-multiplications of nearly equivalent baselines of
types $\{m,n\}$ and $\{m',n'\}$, we get an effective weight:

@@

\caption{Pairplots of the top contributing baseline pairs in three arrays. Plotted properties are optimal time offset $\Delta t_{\boldsymbol{bb'}}$ (in hours),  effective weight $\widetilde{\Theta}_{\boldsymbol{bb'}}$, and baseline length $\bar{b}$ (in meters). \textbf{Shown on the diagonal are histograms of given quantities for the different arrays. On the off-diagonals we show the the scatter plots of the three quantities against each other.} Only those points with $\widetilde{\Theta}_{\boldsymbol{bb'}}>0.01$ are shown (the weight for the top class of equivalent pair is normalized to 1). Scatter points of the three arrays overlap, in the order indicated by the legend. }
\label{fig:pairplot}
\end{figure}

In Fig. \ref{fig:pairplot} we present pair-distributions of three different properties for baselines that contribute well to the sensitivity ($\widetilde{\Theta}_{\boldsymbol{bb'}}>0.01$, where again the weight for the top class of equivalent pair is normalized to 1). The properties shown are effective weight $\widetilde{\Theta}_{\boldsymbol{bb'}}$, optimal time offset $\Delta t_{\boldsymbol{bb'}}$, and average baseline length $\bar{b}$. \textbf{We show the histograms on the diagonals and scatter plots on the off-diagonals.} We only show three of the mentioned arrays for visual clarity. The other two results are qualitatively similar. We point out some key features:
\begin{itemize}

@@

The dashed lines represent the values when only the equivalent baseline-pairs are used. The small and unfractured HERA-37, with no gap (like in HERA-350) or short nearly equivalent baselines (like in PAPER 128), will not benefit much from the nearly equivalent baselines. Near-equivalent baselines make an increasingly significant contribution to the total sensitivity as the number of array elements are increased. PAPER-128 is designed with highly redundant nearly equivalent baselines, and thus  these baselines start contributing at higher $\widetilde{\Theta}_{min}$. \textbf{Incidentally, as we have seen in the scatter plot of $\widetilde{\Theta}_{\boldsymbol{bb'}}$ vs. $\bar{b}$ in Fig. \ref{fig:pairplot}, these are also the shorter baselines contributing to lower $k$-modes.} The fractured HERA configurations will benefit even more from nearly equivalent baselines at low $\widetilde{\Theta}_{min}$ due to the presence of more classes of such pairs. \textbf{However, compared to PAPER, HERA's boost in sensitivity mostly attributes to the longer baselines and thus high $k$-modes (see Fig. \ref{fig:pairplot}). } The stepwise pattern seen in almost all cases are characteristic of a regular grid; as we step to lower $\widetilde{\Theta}_{\boldsymbol{bb'}}$ large groups of baseline pair classes get included in ``batch''. Thus the ``unfractured'' HERA-37, with more regularity in its antenna configurations than the larger counterparts, exhibit more step patterns in fractional sensitivity. 
%\begin{centered}
\begin{figure}[H]
\includegraphics[width=\linewidth]{osens}
\caption{Fractional sensitivity of redundant arrays as a function of the minimum effective weight. Dashed lines represent when only equivalent baseline pairs are used, while solid lines indicate use of both the equivalent and nearly equivalent baselines are used. The vertical axis is normalized independently for each array to the total sensitivity when all baseline pairs, equivalent and near-equivalent, are used.}
\label{fig:osens}
\end{figure}
%\end{centered}
